---
title: "Tornado Events are Most Harmful to Population Health and Economy" 
author: "Denis Levert"
date: "September 11, 2020"
output: html_document
---
## Synopsis

 The intent of this analysis is to obtain the data from NOAA's National
 Weather Center's website. Then to process the data into a usable format.
 The data has been analyzed from two perspectives;
 
 1) The event with the largest impact on peoples health.
 2) The event with the largest impact on economy.
 
 the data clearly indicates that tornadoes have the largest impact on on
 both health and economy.
 
## Data Processing
The first thing is to import the data into R and set up the variables in the
correct class to use in analysis. Data was read in, then formatted the event 
type field to remove spaces, the dates and times into proper format,
The appropriate variables set to factors for proper analysis.

```{r, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
# FUNCTION TO IMPORT DATA AND FORMAT CERTAIN VARIABLES====================
     require(lubridate)
     wd <- read.csv("repdata-data-StormData.csv.bz2")
#FORMAT DATE
     wd$BGN_DATE <- sub(" .*", "", wd$BGN_DATE)
     wd$BGN_DATE <- mdy(wd$BGN_DATE)
#FORMAT TIME
     d <- paste(wd$BGN_DATE,wd$BGN_TIME, wd$TIME_ZONE)
     d <- parse_date_time(d,"%Y-%m-%d %H:%M:%S",
                          truncated = TRUE)
     wd$BGN_TIME <- d
#FORMAT FACTORS
     wd$COUNTYNAME <- as.factor(wd$COUNTYNAME)
     wd$STATE <- as.factor(wd$STATE)
     wd$EVTYPE <- as.factor(wd$EVTYPE)
     wd <<- wd
```
## Results
     
### Which type of events are most harmful with respect to health?
 
The following code is meant to summarize the data by injuries and deaths,
then add them together for a total amount of injuries indicating the effect
on peoples health.
```{r first graph, echo=TRUE, message=FALSE, warning=FALSE}
        require(dplyr)
        require(lattice)
# SUBSET DATA FOR PLOT
        sub <- wd %>% group_by(EVTYPE) %>%
                summarise(fatal = sum(FATALITIES),
                          inj = sum(INJURIES),
                          totinj = sum(FATALITIES) + sum(INJURIES))
        sub <- filter(sub, fatal > 1000 | inj > 1000)
        sub <- arrange(sub, desc(fatal), desc(inj))
# OBTAIN STATS
        a <- max(sub$totinj)
        a <- subset(sub, sub$totinj == a)
        a <- paste("The most harmful event type is", a$EVTYPE , " with", 
                   a$fatal , "fatalities and,", a$inj, "injuries")
# PLOT RESULTS
        xyplot(sub$EVTYPE ~ sub$totinj, sub, 
               col = c("blue"), pch = 19, cex = 1.5, grid = TRUE,
               main = "Total Fatalities & Injuries by Event Type",
               xlab = "Total injuries & deaths",
               ylab = "Event Type")
        print(a)
```

The graph clearly shows that Tornados cause the most injuries to Americans than
any other type of weather event.

## Which type of events have the greatest economic consequences?

The following code will subset the data and total the economic impact of 
events. It will chart the total by event type. The chart clearly indicates
that tornadoes have the largest impact on ecomincs.

```{r economic, echo=TRUE, message=FALSE, warning=FALSE}
# SUBSET DATA FOR PLOT
        require(dplyr)
        require(lattice)
        sub2 <- wd %>% group_by(EVTYPE) %>%
                summarise(pd = sum(PROPDMG), cd = sum(CROPDMG),
                          td = sum(PROPDMG)+sum(CROPDMG))
        sub2 <- filter(sub2,td > 250000)
# OBTAIN STATS
        b <- max(sub2$td)
        b <- subset(sub2, sub2$td == b)
        b <- paste("The most economic impactful weather event is ", b$EVTYPE,
                          " with a property damage of $", 
                   format(b$pd, nsmall = 0, big.mark = ","),
                   " a crop damage of $", 
                   format(b$cd, nsmall = 0, big.mark = ","),
                   " and a total of $",
                   format(b$td, nsmall = 0, big.mark = ","), sep = "")
# PLOT RESULTS
        xyplot(sub2$EVTYPE ~ sub2$td/1000000, sub2, 
               grid = TRUE, pch = 19, cex = 1.5, col = "red",
               main = "Property and Crop Damage vs. Event Type",
               xlab = "Damage Value $M",
               ylab = "Event Type")
        print(b)
```
Tornadoes have the largest financial impact compared to other weather events.


